---
title: "Simulate Admixed Populations with `bnpsd`"
author: "Alejandro Ochoa and John D. Storey"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: bnpsd.bib
vignette: >
  %\VignetteIndexEntry{Simulate Genotypes from the BN-PSD Admixture Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Weird way to include these definitions for HTML Markdown -->
$$
\newcommand{\Fst}{F_{\text{ST}}}
\newcommand{\ft}[1][j]{f_{#1}^T}
$$

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `bnpsd` package simulates the genotypes of an admixed population.
In the PSD model [@pritchard_inference_2000], admixed individuals draw their alleles with individual-specific probabilities (admixture proportions) from $K$ intermediate subpopulations.
We impose the BN model [@balding_method_1995] to the intermediate subpopulation allele frequency, which thus evolve independently with subpopulation-specific inbreeding coefficients ($\Fst$ values) from a common ancestral population $T$.
The kinship coefficients and generalized $\Fst$ of the admixed individuals were derived in recent work [@Ochoa083915].
A simulated admixed population was used to benchmark kinship and $\Fst$ estimators in the accompanying paper [@Ochoa083923].
Here we briefly summarize the notation and intuition behind the key parameters (see [@Ochoa083915] for precise definitions).

## The BN-PSD population structure

The population structure determines how individuals are related to each other.
The key parameters are the inbreeding coefficients of the intermediate subpopulations ($\ft[S_u]$ below) and the admixture proportions of each individual for each subpopulation ($q_{ju}$), which are treated as fixed variables.

Each intermediate subpopulation $S_u$ ($u \in \{1, ..., K\}$) evolved independently from a shared ancestral population $T$ with an inbreeding coefficient denoted by $\ft[S_u]$.
Each admixed individual $j \in \{1, ..., n\}$ draws each allele from $S_u$ with probability given by the admixture proportion $q_{ju}$ ($\sum_{u=1}^K q_{ju} = 1 \forall j$).
In this case the coancestry coefficients $\theta^T_{jk}$ between individuals $j,k$ (including $j=k$ case) and the $\Fst$ of the admixed individuals are given by:
$$
\theta^T_{jk} = \sum_{u=1}^K q_{ju} q_{ku} \ft[S_u],
\quad \quad
\Fst = \sum_{j=1}^n \sum_{u=1}^K w_j q_{ju}^2 \ft[S_u],
$$
where $0 < w_j < 1, \sum_{j=1}^n w_j = 1$ are user-defined weights for individuals (default $w_j=\frac{1}{n} \forall j$).
Note $\theta^T_{jk}$ equals the kinship coefficient for $j \ne k$ and the inbreeding coefficient for $j=k$.

The bias coefficient $s$ is defined by
$$
s = \frac{\bar{\theta}^T}{\Fst}
$$
where $\bar{\theta}^T = \sum_{j=1}^n \sum_{k=1}^n w_j w_k \theta_{jk}^T$.
This $0 < s \le 1$ approximates the proportional bias of $\Fst$ estimators that assume independent subpopulations, and one `bnpsd` function below fits its parameters to yield a desired $s$.

## Random allele frequencies and genotypes

This section details the distributions of the allele frequencies and genotypes of the various populations or individuals of the BN-PSD model.

Every biallelic locus $i$ in the ancestral population $T$ has an ancestral reference allele frequency denoted by $p_i^T$.
By default the `bnpsd` code draws
$$
p_i^T \sim \text{Uniform}(a,b)
$$
with $a=0.01, b=0.5$, but the code accepts $p_i^T$ from arbitrary distributions (see below).

The distribution of the allele frequency at locus $i$ in subpopulation $S_u$, denoted by $p_i^{S_u}$, is the BN distribution:
$$
p_i^{S_u} | T \sim \text{Beta} \left( \nu_s p_i^T, \nu_s \left( 1-p_i^T \right) \right),
$$
where $\nu_s = \frac{1}{\ft[S_u]}-1$.
Allele frequencies for different loci and different subpopulations ($S_u,S_v, u \ne v$) are drawn independently.

Each admixed individual $j$ at each locus $i$ draws alleles from a mixture of Bernoulli distributions from each intermediate subpopulation, which is mathematically equivalent to assigning what we call "individual-specific allele frequencies" (IAFs) $\pi_{ij}$ constructed as:
$$
\pi_{ij} = \sum_{u=1}^K p_i^{S_u} q_{ju}.
$$
The unphased genotype $x_{ij}$ (encoded to count the number of reference alleles) is drawn as:
$$
x_{ij}|\pi_{ij} \sim \text{Binomial}(2, \pi_{ij}).
$$

# Simulation examples

## Population structure: 1D geography

Let's generate the same population structure used in the simulation of [@Ochoa083923].
```{r}
library(RColorBrewer) # for nice colors
library(popkin) # for visualizing coancestry matrix with plot_popkin
library(bnpsd)

# dimensions of data/model
n <- 100 # number of individuals (NOTE this is 10x less than in publication!)
k <- 10 # number of intermediate subpops

# define population structure
inbr_subpops <- 1:k # subpopulation FST vector, up to a scalar
bias_coeff <- 0.5 # desired bias coefficient
Fst <- 0.1 # desired FST for the admixed individuals
# admixture proportions from 1D geography
obj <- admix_prop_1d_linear(
    n_ind = n,
    k_subpops = k,
    bias_coeff = bias_coeff,
    coanc_subpops = inbr_subpops,
    fst = Fst
)
admix_proportions <- obj$admix_proportions
# rescaled inbreeding vector for intermediate subpopulations
inbr_subpops <- obj$coanc_subpops

# get pop structure parameters of the admixed individuals
coancestry <- coanc_admix(admix_proportions, inbr_subpops) # the coancestry matrix
# verify that we got the desired FST!
Fst2 <- fst_admix(admix_proportions, inbr_subpops)
Fst2
# this should also equal FST
inbr <- diag(coancestry)
fst(inbr) # popkin version
# verify that we got the desired bias_coeff too!
bias_coeff2 <- mean(coancestry) / Fst
bias_coeff2
```

```{r, fig.width = 4, fig.height = 2, fig.align = 'center'}
# visualize the per-subpopulation inbreeding coefficients (FSTs)
par(mar = c(4, 4, 0, 0) + 0.2) # shrink default margins
col_subpops <- brewer.pal(k, "Paired") # indep. subpop. colors
barplot(inbr_subpops, col = col_subpops, names.arg = 1:k, ylim = c(0,1),
	xlab = 'Subpopulation', ylab = 'Inbreeding coeff.')
```

```{r, fig.width = 6, fig.height = 1.5, fig.align = 'center'}
# visualize the admixture proportions
par(mar = c(1, 4, 0, 0) + 0.2) # shrink default margins
barplot(t(admix_proportions), col = col_subpops, border = NA, space = 0, ylab = 'Admixture prop.')
mtext('Individuals', 1)
```

```{r, fig.width = 4.2, fig.height = 3, fig.align = 'center'}
# Visualize the coancestry matrix using "popkin"!
# set outer margin for axis labels (left and right are non-zero)
par(oma = c(0, 1.5, 0, 3))
# zero inner margin (plus padding) because we have no individual or subpopulation labels
par(mar = c(0, 0, 0, 0) + 0.2)
# now plot!
plot_popkin(coancestry)
```

## Draw random allele frequencies and genotypes

Now let's draw all the random allele frequencies and genotypes from the population structure.
The easiest way is to use `rbnpsd` (the initial `r` is for drawing "random" samples in this and the following functions, in analogy to the `runif`, `rnorm`, `rbeta`, etc. functions from the `stats` R package).

```{r}
m <- 10000 # number of loci in simulation (NOTE this is 30x less than in publication!)
# draw all random Allele Freqs (AFs) and genotypes
# reuse the previous inbr_subpops, admix_proportions
out <- rbnpsd(admix_proportions, inbr_subpops, m)
X <- out$X # genotypes
p_ind <- out$P # IAFs (individual-specific AFs)
p_subpops <- out$B # intermediate AFs
p_anc <- out$Pa # ancestral AFs
```

```{r, fig.width = 4, fig.height = 2, fig.align = 'center'}
# inspect distribution of ancestral AFs (~ Uniform(0.01, 0.5))
par(mar = c(4, 4, 0, 0) + 0.2) # shrink default margins for these figures
hist(p_anc, xlab = 'Ancestral AF', main = '', xlim = c(0, 1))
# distribution of intermediate population AFs
# (all subpopulations combined)
# (will be more dispersed toward 0 and 1 than ancestral AFs)
hist(p_subpops, xlab = 'Intermediate Subpopulation AF', main = '', xlim = c(0, 1))
# distribution of IAFs (admixed individuals)
# (admixture reduces differentiation, so these resemble ancestral AFs a bit more)
hist(p_ind, xlab = 'Individual-specific AF', main = '', xlim = c(0, 1))
# genotype distribution of admixed individuals
barplot(table(X), xlab = 'Genotypes', ylab = 'Frequency', col = 'white')
```

Lastly, let's verify that the correlation structure of the genotypes matches the theoretical coancestry matrix we constructed earlier.
For this we use the `popkin` function of the package with the same name.

```{r, fig.width = 6, fig.height = 2.8, fig.align = 'center'}
# for best estimates, group individuals into subpopulations using the geography
# this averages more individuals in estimating the minimum kinship
subpops <- ceiling( (1:n) / n * k )
table(subpops) # got k=10 subpops with 100 individuals each
# now estimate kinship using popkin
kinship_estimate <- popkin(X, subpops)
# replace diagonal with inbreeding coeffs. to match coancestry matrix
coancestry_estimate <- inbr_diag(kinship_estimate)

# Visualize the coancestry matrix using "popkin"!
# set outer margin for axis labels (left and right are non-zero)
par(oma = c(0, 1.5, 0, 3))
# increase inner top margin for panel titles
par(mar = c(0, 0, 2.5, 0) + 0.2)
# now plot!
coancestries <- list(coancestry, coancestry_estimate)
titles <- c('Truth', 'Estimate')
plot_popkin(coancestries, titles)
```


## Customizing the allele frequency and genotype pipeline

The random variables generated by `rbnpsd` above can also be generated separately using the following functions (where $p$ is the usual variable symbol for allele frequencies):

* `draw_p_anc` (drawn ancestral allele frequencies)
* `draw_p_subpops` (draw independent subpopulation allele frequencies)
* `make_p_ind_admix` (construct individual-specific allele frequencies under admixture model)
* `rgeno` (random GENOtypes)

Here is the step-by-step procedure for drawing AFs and genotypes in `rbnpsd`:

```{r}
# reuse the previous m, inbr_subpops, admix_proportions
p_anc <- draw_p_anc(m) # draw ancestral AFs
p_subpops <- draw_p_subpops(p_anc, inbr_subpops) # draw intermediate AFs
p_ind <- make_p_ind_admix(p_subpops, admix_proportions) # draw IAFs (individual-specific AFs)
X <- rgeno(p_ind) # draw genotypes
```

We provide functions for these separate steps to allow for more flexibility.
For example, the ancestral allele frequencies could be drawn from a Symmetric Beta:

```{r, fig.width = 4, fig.height = 2, fig.align = 'center'}
alpha <- 1/2 # this increases rare alleles
p_anc <- rbeta(m, alpha, alpha)
par(mar = c(4, 4, 0, 0) + 0.2) # shrink default margins for these figures
hist(p_anc, xlab = 'Ancestral AF', main = '', xlim = c(0, 1))
```

You could also draw genotypes from the ancestral population or the intermediate populations:

```{r}
# draw genotypes for one individual from the ancestral population
# use "cbind" to turn the vector p_anc into the column matrix "rgeno" expects
X_anc <- rgeno(cbind(p_anc))
# returns a column matrix:
dim(X_anc)
# draw genotypes from intermediate populations
# draws one individual per intermediate population
X_subpops <- rgeno(p_subpops)
```

## Low-memory genotype simulation algorithm

If you desire to simulate a very large number of individuals ($n$) and loci ($m$), you might run out of memory while running `rbnpsd`.
Memory consumption is reduced by passing the `lowMem = TRUE` option to `rbnpsd`, which draws the genotypes `X` directly from the subpopulation AF matrix `p_subpops` and `admix_proportions`, without storing the whole IAF matrix `p_ind` at any given time.
However, this algorithm is much slower than the default one (`lowMem = FALSE`).

```{r}
out <- rbnpsd(admix_proportions, inbr_subpops, m, lowMem = TRUE)
X <- out$X # genotypes
p_subpops <- out$B # intermediate AFs
p_anc <- out$Pa # ancestral AFs
# NOTE: out$P is not computed in this mode!
```

The same option exists for `rgeno`, which instead of accepting the IAF matrix `p_ind` as input requires both `p_subpops` and `admix_proportions` as above:

```{r}
X <- rgeno(p_subpops, admix_proportions, lowMem = TRUE)
```




# Additional population structures

Here we show examples for functions that create admixture matrices for various simple population structures.
The admixture scenarios implemented in `bnpsd` are generated by these functions:

* `admix_prop_1d_linear`: Linear 1D geography
* `admix_prop_1d_circular`: Circular 1D geography
* `admix_prop_indep_subpops`: Independent Subpopulations

The first example (above) was `admix_prop_1d_linear`, the rest follow.


## Circular 1D geography

This is a twist on the earlier 1D geography where subpopulations and individuals are placed on a circumference, so random walks wrap around and the appropriate density is the Von Misses distribution.

Let's generate an analogous population structure to the original "linear" example.
```{r}
# reuse earlier (n,k) dimensions
n <- 100 # number of individuals
k <- 10 # number of intermediate subpops

# define population structure
inbr_subpops <- 1:k # subpopulation FST vector, up to a scalar
bias_coeff <- 0.5 # desired bias coefficient
Fst <- 0.1 # desired FST for the admixed individuals
# admixture proportions from *circular* 1D geography
obj <- admix_prop_1d_circular(
    n_ind = n,
    k_subpops = k,
    bias_coeff = bias_coeff,
    coanc_subpops = inbr_subpops,
    fst = Fst
)
admix_proportions <- obj$admix_proportions
inbr_subpops <- obj$coanc_subpops

# get pop structure parameters of the admixed individuals
coancestry <- coanc_admix(admix_proportions, inbr_subpops) # the coancestry matrix
# verify that we got the desired FST!
fst_admix(admix_proportions, inbr_subpops)

# verify that we got the desired bias_coeff too!
mean(coancestry) / Fst
```

```{r, fig.width = 4, fig.height = 1.2, fig.align = 'center'}
# visualize the per-subpopulation inbreeding coefficients (FSTs)
par(mar = c(2.5, 2.5, 0.3, 0) + 0.2, lab = c(2, 1, 7), mgp = c(1.5, 0.5, 0)) # tweak margins/etc
col_subpops <- brewer.pal(k, "Paired") # indep. subpop. colors
barplot(inbr_subpops, col = col_subpops, names.arg = colnames(admix_proportions), xlab = 'Subpopulation', ylab = 'Inbr')
```

```{r, fig.width = 4, fig.height = 1, fig.align = 'center'}
# visualize the admixture proportions
par(mar = c(1, 4, 0.4, 0) + 0.2, lab = c(2, 2, 7)) # tweak margins/etc
barplot(t(admix_proportions), col = col_subpops, border = NA, space = 0, ylab = 'Admix prop')
mtext('Individuals', 1)
```

```{r, fig.width = 3.1, fig.height = 2, fig.align = 'center'}
# Visualize the coancestry matrix using "popkin"!
par(oma = c(0, 1.5, 0, 3), mar = c(0, 0, 0.4, 0) + 0.2) # tweak margins/etc
plot_popkin(coancestry, leg_n = 3)
```



## Independent subpopulations

The independent subpopulations model, where individuals are actually unadmixed, is the most trivial form of the BN-PSD admixture model.

```{r}
# define population structure
# we'll have k=3 subpopulations, each with these sizes:
k <- 3
n1 <- 100
n2 <- 50
n3 <- 20
# here's the labels (for simplicity, list all individuals of S1 first, then S2, then S3)
labs <- c(
    rep.int('S1', n1),
    rep.int('S2', n2),
    rep.int('S3', n3)
)
# data dimensions infered from labs:
length(labs) # number of individuals "n"
length(unique(labs)) # number of subpopulations "k"

# desired admixture matrix
admix_proportions <- admix_prop_indep_subpops(labs)
# got a numeric matrix with a single 1 value per row
# (denoting the sole subpopulation from which each individual draws its ancestry)
head(admix_proportions, 2)

# construct the intermediate subpopulation FST vector
# NOTE fst is a function in the `popkin` package
Fst <- 0.2 # the desired final FST
inbr_subpops <- 1:k # subpopulation FST vector, unnormalized so far
inbr_subpops <- inbr_subpops / fst(inbr_subpops) * Fst # normalized to have the desired FST
fst(inbr_subpops) # verify FST for the intermediate subpopulations

# get coancestry of the admixed individuals
coancestry <- coanc_admix(admix_proportions, inbr_subpops) # the coancestry matrix
# before getting FST for individuals, weigh then inversely proportional to subpop sizes
weights <- weights_subpops(labs) # function from `popkin` package
# verify FST for individuals (same as for intermediate subpops for this pop structure)
fst_admix(admix_proportions, inbr_subpops, weights)
```

```{r, fig.width = 4, fig.height = 1.2, fig.align = 'center'}
# visualize the per-subpopulation inbreeding coefficients (FSTs)
par(mar = c(2.5, 2.5, 0, 0) + 0.2, lab = c(2, 1, 7), mgp = c(1.5, 0.5, 0)) # tweak margins/etc
col_subpops <- brewer.pal(k, "Paired") # indep. subpop. colors
barplot(inbr_subpops, col = col_subpops, names.arg = colnames(admix_proportions), xlab = 'Subpopulation', ylab = 'Inbr')
```

```{r, fig.width = 4, fig.height = 1, fig.align = 'center'}
# visualize the admixture proportions
par(mar = c(1, 4, 0.4, 0) + 0.2, lab = c(2, 2, 7)) # tweak margins/etc
barplot(t(admix_proportions), col = col_subpops, border = NA, space = 0, ylab = 'Admix prop')
mtext('Individuals', 1)
```

```{r, fig.width = 3.1, fig.height = 2, fig.align = 'center'}
# Visualize the coancestry matrix using "popkin"!
par(oma = c(0, 1.5, 0, 3), mar = c(0, 0, 0.4, 0) + 0.2) # tweak margins/etc
plot_popkin(coancestry, leg_n = 3)
```

# References
